---
title: "Genomic Analysis"
output: html_document
---
# Format CNA data sourced from cBio
```{r, message = FALSE, warning = FALSE}
# TCGA
TCGA <- read.table("TCGA_data.txt", check.names = FALSE, sep = "\t")
TCGA <- as.data.frame(TCGA)
# Taylor
Taylor <- read.table("Taylor_data.txt", check.names = FALSE, sep = "\t")
Taylor <- as.data.frame(Taylor)
# Baca
Baca <- read.table("Baca_data.txt", check.names = FALSE, sep = "\t")
Baca <- as.data.frame(Baca)
# Abida
Abida <- read.table("Abida_data.txt", check.names = FALSE, sep = "\t")
Abida <- as.data.frame(Abida)
```
Combined primary datasets
```{r}
primary <- merge.data.frame(Taylor, TCGA)
primary <- merge.data.frame(primary, Baca)
met <- Abida
```
Filter out genes missing in both primary and metastatic settings
```{r}
# Obtain genes present in primary and metastatic CNA data
genes <- intersect(primary$Hugo_Symbol, met$Hugo_Symbol)
# Create filtering index for primary dataset
primary.index <- c()
for (i in 1:nrow(primary)){
  if (sum(primary$Hugo_Symbol[i] == genes) == 1)
  {primary.index[i] = TRUE}
  if (sum(primary$Hugo_Symbol[i] == genes) == 0)
  {primary.index[i] = FALSE}
}
# Create filtering index for metastatic dataset
met.index <- c()
for (i in 1:nrow(met)){
  if (sum(met$Hugo_Symbol[i] == genes) == 1)
  {met.index[i] = TRUE}
  if (sum(met$Hugo_Symbol[i] == genes) == 0)
  {met.index[i] = FALSE}
}
# Filter primary and metastatic genes. The number of rows for primary and metastatic datasets should now be equal.
primary <- primary[primary.index,]
primary <- primary[duplicated(primary)==FALSE,]
met <- met[met.index,]
met <- met[duplicated(met)==FALSE,]
```
Source analysis scripts
```{r, message=FALSE}
source("analysis scripts.R")
```
Subset coloss and WT patients
```{r}
# Primary
primary.coloss <- subtype_subset(subtype_subset(primary, c("MAP3K7", "loss")), c("CHD1", "loss"))
primary.WT <- subtype_subset(subtype_subset(primary, c("MAP3K7", "WT/gain")), c("CHD1", "WT/gain"))
# Metastatic
met.coloss <- subtype_subset(subtype_subset(met, c("MAP3K7", "loss")), c("CHD1", "loss"))
met.WT <- subtype_subset(subtype_subset(met, c("MAP3K7", "WT/gain")), c("CHD1", "WT/gain"))
```
# Compute subtype-specific metastatic regulator enrichment
```{r}
del.enrichment <- concurrent.loss.data(met.coloss, primary.coloss)
amp.enrichment <- concurrent.loss.data(met.coloss, primary.coloss)
```
# Compute alteration rates
Compute deletion rates in subtype patients
```{r}
# Primary
primary.coloss.del <- lossRank(primary.coloss)
primary.coloss.del <- primary.coloss.del[,c(1,5)]
colnames(primary.coloss.del)[2] <- "Deletion rate: Primary coloss patients"
# Metastatic
met.coloss.del <- lossRank(met.coloss)
met.coloss.del <- met.coloss.del[,c(1,5)]
colnames(met.coloss.del)[2] <- "Deletion rate: Metastatic coloss patients"
```
Compute deletion rates in WT patients
```{r}
# Primary
primary.WT.del <- lossRank(primary.WT)
primary.WT.del <- primary.WT.del[,c(1,5)]
colnames(primary.WT.del)[2] <- "Deletion rate: Primary WT patients"
# Metastatic
met.WT.del <- lossRank(met.WT)
met.WT.del <- met.WT.del[,c(1,5)]
colnames(met.WT.del)[2] <- "Deletion rate: Metastatic WT patients"
```
Compute amplification rates in subtype patients
```{r}
# Primary
primary.coloss.amp <- gainRank(primary.coloss)
primary.coloss.amp <- primary.coloss.amp[,c(1,5)]
colnames(primary.coloss.amp)[2] <- "Amplification rate: Primary coloss patients"
# Metastatic
met.coloss.amp <- gainRank(met.coloss)
met.coloss.amp <- met.coloss.amp[,c(1,5)]
colnames(met.coloss.amp)[2] <- "Amplification rate: Metastatic coloss patients"
```
Compute amplification rates in WT patients
```{r}
# Primary
primary.WT.amp <- gainRank(primary.WT)
primary.WT.amp <- primary.WT.amp[,c(1,5)]
colnames(primary.WT.amp)[2] <- "Amplification rate: Primary WT patients"
# Metastatic
met.WT.amp <- gainRank(met.WT)
met.WT.amp <- met.WT.amp[,c(1,5)]
colnames(met.WT.amp)[2] <- "Amplification rate: Metastatic WT patients"
```
# Filter subtype-specific metastatic regulators
### Create data frames for filtering
Deletions
```{r}
deletions <- merge.data.frame(primary.WT.del, primary.coloss.del)
deletions <- merge.data.frame(deletions, met.WT.del)
deletions <- merge.data.frame(deletions, met.coloss.del)
```
Amplifications
```{r}
amplifications <- merge.data.frame(primary.WT.amp, primary.coloss.amp)
amplifications <- merge.data.frame(amplifications, met.WT.amp)
amplifications <- merge.data.frame(amplifications, met.coloss.amp)
```
### Constraint 1 (C1): Alteration rate is greater in the primary setting (gene is co-altered in primary patients)
Deletion rate greater in primary subtype
```{r}
index.del.C1 <- (deletions$`Deletion rate: Primary coloss patients` > deletions$`Deletion rate: Primary WT patients`)
del.C1.genes <- deletions$Hugo_Symbol[index.del.C1]
```
Amplification rate greater in primary subtype
```{r}
index.amp.C1 <- (amplifications$`Amplification rate: Primary coloss patients` > amplifications$`Amplification rate: Primary WT patients`)
amp.C1.genes <- amplifications$Hugo_Symbol[index.amp.C1]
```
### Constraint 2 (C2): Alteration rate in WT metastatic patients ~ alteration rate in WT primary patients (gene is not under metastatic selection pressure)
Determine reasonable cutoff for C2
```{r}
library(ggplot2)
# Test genes captured at different effect sizes and report as table
C2.table <- data.frame()
for (i in seq(.2, 1e-5, length.out = 20)){
  index.del.C2 <- (abs(deletions$`Deletion rate: Metastatic WT patients`-deletions$`Deletion rate: Primary WT patients`) <= i)
  del.C2.genes <- deletions$Hugo_Symbol[index.del.C2]
  index.amp.C2 <- ((abs(amplifications$`Amplification rate: Metastatic WT patients`- amplifications$`Amplification rate: Primary WT patients`)) <= i)
  amp.C2.genes <- amplifications$Hugo_Symbol[index.amp.C2]
  del.genes <- intersect(del.C1.genes, del.C2.genes)
  amp.genes <- intersect(amp.C1.genes, amp.C2.genes)
  genes.captured <- length(del.genes) + length(amp.genes)
  C2.table[nrow(C2.table)+1,c(1,2)] <- c(i, genes.captured)
}
colnames(C2.table) <- c("WT Alteration Ratio Difference: |Met - Primary|", "Genes Remaining")
library(DT)
ggplot(data = C2.table, aes(`WT Alteration Ratio Difference: |Met - Primary|`, `Genes Remaining`)) + geom_point() + xlab("Constraint 2 Cutoff")
print("A C2 cutoff of <= 2% captures ~3000 genes that remain for analysis as subtype-specific metastatic regulators")
```
Apply C2 cutoff
```{r}
# Set cutoff
cutoff <- 0.02
# Filter deletions
index.del.C2 <- (abs(deletions$`Deletion rate: Metastatic WT patients`-deletions$`Deletion rate: Primary WT patients`) <= cutoff)
del.C2.genes <- deletions$Hugo_Symbol[index.del.C2]
# Filter ampplifications
index.amp.C2 <- ((abs(amplifications$`Amplification rate: Metastatic WT patients`- amplifications$`Amplification rate: Primary WT patients`)) <= cutoff)
amp.C2.genes <- amplifications$Hugo_Symbol[index.amp.C2]
```
# Report ranked alterations
Assign alteration status
```{r}
filtered.deletions <- intersect(del.C1.genes, del.C2.genes)
filtered.amplifications <- intersect(amp.C1.genes, amp.C2.genes)
deleted <- setdiff(filtered.deletions, filtered.amplifications)
amplified <- setdiff(filtered.amplifications, filtered.deletions)
both <- intersect(filtered.deletions, filtered.amplifications)
```
Construct summary table
```{r}
# Construct alteration table for genes that are deleted
if (length(deleted) > 0){
  deleted <- as.data.frame(deleted)
  deleted[,2] <- "Deletion"
  for (i in (1:nrow(deleted))) {
    index1 <- which(met.coloss.del$Hugo_Symbol == deleted[i,1])
    index2 <- which(primary.coloss.del$Hugo_Symbol == deleted[i,1])
    index3 <- which(del.enrichment$Hugo_Symbol == deleted[i,1])
    deleted[i,3] <- met.coloss.del[index1, 2] - primary.coloss.del[index2, 2]
    deleted[i,4] <- del.enrichment[index3, 9]
  }
colnames(deleted) <- c("Hugo Symbol", "Alteration", "Subtype Alteration Ratio Difference (Met - Primary)","pval")
}
# Construct alteration table for genes that are amplified
if (length(amplified) >0){
  amplified <- as.data.frame(amplified)
  amplified[,2] <- "Amplification"
  for (i in (1:nrow(amplified))) {
    index1 <- which(met.coloss.amp$`Hugo_Symbol` == amplified[i,1])
    index2 <- which(primary.coloss.amp$Hugo_Symbol == amplified[i,1])
    index3 <- which(amp.enrichment$Hugo_Symbol == amplified[i,1])
    amplified[i,3] <- met.coloss.amp[index1, 2] - primary.coloss.amp[index2, 2]
    amplified[i,4] <- amp.enrichment[index3, 9]
  }
colnames(amplified) <- c("Hugo Symbol", "Alteration", "Subtype Alteration Ratio Difference (Met - Primary)","pval")
}
# Construct alteration table for genes that are deleted and amplified. Report the average change in alteration and the combined p-val.
if (length(both) >0){
  library(metap)
  both <- as.data.frame(both)
  both[,2] <- "Amplification & Deletion"
  for (i in (1:nrow(both))) {
    index1a <- which(met.coloss.del$`Hugo_Symbol` == both[i,1])
    index1b <- which(met.coloss.amp$`Hugo_Symbol` == both[i,1])
    index2a <- which(primary.coloss.del$`Hugo_Symbol` == both[i,1])
    index2b <- which(primary.coloss.amp$Hugo_Symbol == both[i,1])
    index3a <- which(del.enrichment$Hugo_Symbol == deleted[i,1])
    index3b <- which(amp.enrichment$Hugo_Symbol == both[i,1])
    both[i,3] <- (((met.coloss.del[index1a, 2] - primary.coloss.del[index2a, 2]) + 
                     (met.coloss.amp[index1b, 2] - primary.coloss.amp[index2b, 2])) / 2)
    both[i,4] <- sumlog(c(del.enrichment[index3a, 9], amp.enrichment[index3b, 9]))$p
  }
  #both[,4] <- p.adjust(both[,4], method = "fdr")
colnames(both) <- c("Hugo Symbol", "Alteration", "Average Subtype Alteration Ratio Difference (Met - Primary)","pval")
}
# Combine alteration tables
if (length(both) == 0){
  if ((length(deleted) > 0) & (length(amplified) > 0)){
  summary <- rbind(deleted, amplified)
}
  if ((length(deleted) > 0) & (length(amplified) == 0)){
  summary <- (deleted)
}
  if ((length(deleted) == 0) & (length(amplified) > 0)){
  summary <- (amplified)
  }
}
if (length(both) > 0){
    if ((length(deleted) > 0) & (length(amplified) > 0)){
  summary <- rbind(deleted, amplified)
  summary <- rbind(summary, both)
}
  if ((length(deleted) > 0) & (length(amplified) == 0)){
  summary <- (deleted)
  summary <- rbind(summary, both)

}
  if ((length(deleted) == 0) & (length(amplified) > 0)){
  summary <- (amplified)
  summary <- rbind(summary, both)
  }
}
library(DT)
datatable(summary)
```
p-value distribution
```{r}
metastatic <- summary$pval
hist(metastatic, breaks = seq(0, 1, by = 0.05),
     xlab = "p value", ylab = "Number of genes", main = "Met:Primary Subtype Enrichment")
```
Order summary table by difference in alteration ratio
```{r}
library(ggplot2)
library(ggrepel)
# Order hits by increasing alteration rate in metastatic subtype
summary <- summary[order(summary$`Subtype Alteration Ratio Difference (Met - Primary)`, decreasing = TRUE),]
summary[,5] <- c(1:nrow(summary))
colnames(summary)[5] <- "Gene Rank"
# Add annotations
annotations <- Bands(summary$`Hugo Symbol`)
summary <- cbind(summary, annotations$Band)
colnames(summary)[6] <- "Cytogenetic Band"
# Report summary
datatable(summary)
# Plot ranked summary data
ggplot(data = summary, mapping = aes(`Gene Rank`, `Subtype Alteration Ratio Difference (Met - Primary)`, label = `Hugo Symbol`)) + geom_point() + ggtitle("Alteration Frequency Difference: Met - Primary") + ylim(-.4,.4) + geom_text_repel(data = subset(summary, `Gene Rank` <= 2 | `Gene Rank` >= 3044), nudge_x = 0, nudge_y = -.165)
```
